Nonlinear Dynamical Tides in White Dwarf Binaries

### 2019-04-26 writeups/paper.tex
check in new abstract?

\begin{abstract}
    In white dwarf binaries tidal torques are expected to raise a train of
    internal gravity waves in the white dwarf interior. These
    outward-propagating waves can then grow to be nonlinear and break via
    hydrodynamical instabilities, transfering energy and angular momentum from
    the binary orbit to the white dwarf. We perform 2D numerical simulations of
    nonlinear wave breaking of outwards-propagating internal gravity waves in an
    incompressible, isothermal atmosphere with an exponentially decaying density
    stratification. We find that tidal synchronization, after an initial
    transient phase, proceeds inwards from the surface in thin layers. We argue
    the thickness of these layers is limited by the Kelvin-Helmholtz
    Instability. We provide simple analytical formulae for the location and
    thickness of tidal synchronization over time that are in good agreement with
    our simulations.
\end{abstract}

\title{Tidal Heating via Internal Gravity Waves in White Dwarf Binaries}
\author{Authors}
\date{\today}
\maketitle

@@

We consider a two-dimensional incompressible, isothermal fluid, representative
of degenerate matter in WDs. We neglect temperature variations and assume a
barotropic equation of state as a first approximation. As we are interested in
dynamics far from the center of the WD, we approximate the gravitational field
as uniform. We model the background density stratification as $\rho_0(x, z) =
\rho_0(z) = \rho_0(z=0) e^{-z/H}$ for some reference density $\rho_0(z=0) =
\at{\rho_0(z)}_{z = 0}$.

@@

In the weak stratification limit $k_{0z}H \gg 1$, the solution exhibits the
following characteristics:
\begin{itemize}
    \item The amplitude of the wave grows like $e^{z/2H}$. Thus, the
        linear approximation is violated for sufficiently large $z$.

@@

In developing the results above, we have neglected advective terms $\pd{}{t} \gg
\vec{u} \cdot \vec{\nabla}$. However, since $\abs*{\vec{u}} \propto e^{z/2H}$,
any nonzero amplitude IGW will eventually violate the linearity criterion. This
regime is relevant in practice since the IGW may grow to nonlinear amplitudes
before it reaches the surface of the WD\cite{fullerI,fullerII}. We can coarsely
estimate the height at which this happens by when the Lagrangian displacement
$\vec{\xi}$ of a fluid parcel satisfies
\begin{equation}
    \vec{\xi} \cdot \vec{k} \gtrsim 1.\label{eq:nl}
\end{equation}

@@

An estimate of where these mean flow effects become important can be made using
the wave-induced mean flow\cite{eliassen_palm_cite,sutherland0}:
\begin{equation}
     \ev*{u_x} = \bar{U}_0 \equiv \frac{1}{L_x}
        \int\limits u_x\;\mathrm{d}x = \frac{\ev*{u_xu_z}_x}{c_{g,z}},
\end{equation}
where $\ev*{\dots}_x$ denotes horizontal averaging. Setting $\bar{U}_0 = c_{ph,
x}$ the critical layer condition, we recover $\frac{u_xk_x}{\omega} = \xi_x k_x
\sim 1$, roughly equivalent to the nonlinearity criterion \autoref{eq:nl}.

@@

To study IGW breaking as a weak perturbation grows to nonlinear amplitudes, we
perform direct numerical simulation using Dedalus\cite{dedalus} a spectral
numerical method. Our simulation domain is smaller than our full physical
domain, so we adopt reflection-suppressing boundary conditions. We use periodic
boundary conditions in the $x$ direction and damping layers (described in
\autoref{ss:damping}) in the $z$ direction to damp perturbations exiting the
top and bottom of the domain. We use a Fourier basis in the $x$ direction and
either a Chebyshev or Fourier basis in the $z$ direction; our results were
robust to either set of basis functions. We varied the number of $x, z$ modes
(denoted $N_x, N_z$ respectively), and used $3/2$ dealiasing\cite{boyd}.

The time integration uses a split implicit-explicit first-order scheme where
certain terms are treated implicitly using for- ward Euler, and the remaining
terms are treated explicitly using backward Euler. A third-order, four-stage
DIRK-ERK scheme\cite{ascher} is used with adaptive timesteps taken to be
the advective Courant-Friedrichs-Lewy (CFL) time given by $0.7 \min (\Delta
x / u_x,\Delta z / u_z)$. The minimum is taken over every grid point in
the domain and $\Delta x,\Delta z$ are the grid spacings in the $x$ and $z$
directions respectively.

@@

Denote $L_x, L_z$ to be the physical dimensions of the simulation domain. The
physics of our simulation is fixed by six parameters: $L_x, L_z$, and $k_{0x},
\omega_0, F, \nu$. We describe and choose these parameters below:
\begin{itemize}
    \item $L_z$: Since the wave amplitude grows $\sim e^{z/2H}$, we choose $L_z
        = 10H$ to give $\sim e^4$ apmlitude growth between the damping zones.

    \item $L_x$: We want similar grid spacing $\frac{L_x}{N_x} \sim
        \frac{L_z}{N_z}$. For computational savings, we fix $N_z = 4N_x$.
        Expecting wave breaking to be mostly in the $z$ direction, we use $L_x =
        4H$.

    \item $k_{0x}$: Astrophysical IGWs in stars are generally excited by the $l
        = 2$ component of the tidal potential, for which $k_{\perp} \sim
        \frac{1}{R}$ where $R$ is the radius of the star. To best emulate this,
        we use $k_{0x} = \frac{2\pi}{L_x}$ the smallest permitted wavenumber
        permitted by periodic boundary conditions.

    \item $\omega_0$: We choose $\omega_0$ by evaluating dispersion relation
        $\omega_0\p*{k_{0x}, k_{0z}}$ for some desired $k_{0z}$.

        Astrophysical IGWs also generally satisfy $\omega \ll N$, or
        equivalently $\frac{k_r}{k_{\perp}} \sim k_rR \gg 1$. Since $H \lesssim
        R$, we aim to study $k_{0z}H \gg 1$. However, to satisfy this well is
        computationally expensive: $L_z \gg H$ is required to give waves ample
        room to grow within the simulation domain, but we want $k_{0z}$ to be
        sufficiently separated from the grid spacing $\sim L_z / N_z$ that
        nonlinear effects can be seen in later simulations (see
        \autoref{ss:nl_ns}). Thus, $N_z$ sets how well we can satisfy $k_{0z}H
        \gg 1$; we choose $k_{0z}H = 1$.

    \item $F$: We first choose $F$ forcing strength such that $\vec{\xi} \cdot
        \vec{k} \ll 1$ is satisfied everywhere throughout the domain. This
        becomes a constraint on $F$ by using \autoref{eq:uz_lin} and
        $\abs*{\vec{u}} \propto e^{z/2H}$.

    \item $\nu$: Nonlinear effects transfer wave energy from $\vec{k}_{0}$ to
        smaller scales. Since well-resolved simulations using spectral methods
        have no inherent numerical viscosity, energy will accumulate at grid
        scales in the absence of artificial dissipation. We introduce
        Navier-Stokes viscosity $\nu$ while ensuring our equations of motion
        continue to conserve horizontal momentum (see \autoref{ss:strat_impl}
        for details). When $\vec{\xi} \cdot \vec{k} \ll 1$, we may set $\nu =
        0$.
\end{itemize}

@@

To imitate an infinite fluid using a finite simulation domain, we use periodic
boundary conditions in the $x$ direction and damping layers at the top and
bottom of the $z$ direction. This damps waves that reach the edge of the
simulation domain without inducing nonphysical reflection. Specifically, these
are implemented by introducing linear damping terms to \autoref{se:nl_upsilon}:
\begin{align}
    \md{\Upsilon}{t} + u_z \pd{\ln \rho_0}{z} &=
        -\Gamma(z)\Upsilon,\nonumber\\
    \md{\vec{u}}{t} + \vec{\nabla}P + P\vec{\nabla}\Upsilon + g\hat{z} &=
        - \Gamma(z)\vec{u},\nonumber\\
    \frac{1}{2\tau}\s*{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}} &= \Gamma(z),\label{eq:Gamma}
\end{align}
where $z_B = 0.05L_z, z_T = 0.95L_z$ are the boundaries of the damping zone.
This strongly damps perturbations below $z_B$ and above $z_T$ with damping time
$\tau$, negligibly affects dynamics between $z_B, z_T$ and has transition width
governed by $\Delta z$; we use $\Delta z = 0.025L_z$. This prescription, used in
similar studies\cite{lecoanet_damp}, has the advantage of being smooth,
important for spectral methods. The full implementation of the fluid equations
in Dedalus is described in \autoref{ss:strat_impl}.

\subsection{Linear Regime}\label{ss:lin_ns}

We first test our numerical simulation with a ``linear'' simulation, where
$F$ is chosen such that $\vec{\xi} \cdot \vec{k} \ll 1$ everywhere. We emphasize
that our ``linear'' simulations solve the full nonlinear equations, just in the
weak perturbation limit.

@@

A Navier-Stokes numerical viscosity $\nu$ is used to damp high wavenumbers and
regularize the nonlinear cascade at near grid resolution: $\nu \sim 0.1
\frac{\omega_0}{\abs*{k_{0z}}}\frac{L_z}{2\pi N_z}$ was found to be suitable for
$N_z = 2048$.

@@

\subsection{Numerical Simulation}\label{ss:nl_ns}

@@

Two representative snapshots from our simulation are provided in
TODO after the critical layer has had time to form. We may note that
the critical layer, where $S_{px}$ is absorbed and $\bar{U}_0 = \bar{U}_c$,
travels downwards as predicted.

@@

For the simulation in TODO, we may define $z_c = \argmax_z
\pd{\bar{U}_0}{z}$. Computing $\pd{z_c}{t}$ using the analytical flux
\autoref{eq:fpx_lin} allows comparison to \autoref{eq:zc_anal}, which we exhibit
in TODO\@. We see overall good agreement, though some small
deviation is expected since $S_{px}$ is not perfectly conserved owing to
numerical viscosity (and YUBONOTE $S_{px}$ is misestimated?).

However, recalling \autoref{eq:crit_coeffs}, we only expect complete critical
layer absorption when $\mathrm{Ri} \gg \frac{1}{4}$. We may plot $\mathrm{Ri}$
over the same time interval in TODO\@. We may observe that the
Richardson number is initially decreasing but eventually levels out and
begins to increase again. This corresponds to an increasingly sharp critical
layer transition then subsequently a decreasingly sharp critical layer
transition.

We argue that the Richardson number is bounded from below by viscosity. A
repeated simulation with a larger viscosity is shown in TODO,
where the Richardson number does not go nearly as low.

>>>>

\begin{abstract}
    In sufficiently compact white dwarf binaries, dynamical tides raise a train
    of internal gravity waves that propagate towards the surface. We perform 2D
    numerical simulations of these waves undergoing nonlinear wave breaking in
    an incompressible, isothermal atmosphere. After an initial transient phase,
    we find that these waves induce a sharp transition between a non-rotating
    core and synchronously rotating envelope. We find evidence that the width of
    this transition layer is bound from below by the Kelvin-Helmholtz
    Instability. We provide analytical formulae for absorption and reflection of
    incident waves off the critical layer up to prefactors of order unity. These
    prefactors converge to constant values when artificial dissipation is
    decreased. We provide dimensionless criteria necessary to resolving momentum
    transfer within the critical layer. Finally, we speculate on the application
    of our model to tidal synchronization and heating in astrophysical systems.
\end{abstract}

\title{Nonlinear Dynamical Tides in White Dwarf Binaries}
\author{Authors}
\date{\today}
\maketitle

@@

We consider an incompressible, isothermal fluid, representative of degenerate
matter in WD bulks, in 2D. We neglect temperature variations and assume a
barotropic equation of state as a first approximation. As we are interested in
the behavior of the wave far from the center of the WD, we approximate the
gravitational field as uniform. We model the background density stratification
as $\rho_0(x, z) = \rho_0(z) = \rho_0(z=0) e^{-z/H}$ for some reference density
$\rho_0(z=0) = \at{\rho_0(z)}_{z = 0}$.

@@

In the weak stratification limit $k_{0z}H \gg 1$, the solution exhibits the
following characteristics:
\begin{itemize}
    \item The amplitude of the perturbations grows like $e^{z/2H}$. Thus, the
        linear approximation is violated for sufficiently large $z$.

@@

In developing the results above, we have neglected advective terms $\pd{}{t} \gg
\vec{u} \cdot \vec{\nabla}$. However, since $\abs*{\vec{u}} \propto e^{z/2H}$,
any nonzero amplitude IGW will eventually violate the linearity criterion. This
regime is relevant in practice since the IGW grows to nonlinear amplitudes
before it reaches the surface of the WD\cite{fullerI,fullerII}. We can coarsely
estimate the height at which this happens by $\abs*{\vec{u}}(z) \gtrsim
\frac{\omega}{\abs*{\vec{k}}}$. This corresponds to when the Lagrangian
displacement $\vec{\xi} \equiv \vec{u} / \omega$ satisfies $\vec{\xi} \cdot
\vec{k} \gtrsim 1$, the same criterion used in\cite{fullerI}.

@@

An estimate of where these mean flow effects become important can be made using
the wave-induced mean flow formula\cite{eliassen_palm_cite,sutherland0}:
\begin{equation}
     \ev*{u_x} = \bar{U}_0 \equiv \frac{1}{L_x}
        \int\limits u_x\;\mathrm{d}x = \frac{\ev*{u_xu_z}_x}{c_{g,z}},
\end{equation}
where $\ev*{\dots}x$ denotes horizontal averaging. Setting $\bar{U}_0 = c_{ph,
x}$ the critical layer condition, we can solve that $\vec{\xi} \cdot \vec{k}
= \sqrt{\frac{k_x}{k_z}}$ is where the mean flow attains the horizontal phase
velocity. In an astrophysical context, $N \gg \omega$ and $k_z \gg k_x$. Thus,
wave breaking may occur significantly earlier than assumed in current
astrophysics literature.

@@

To study IGW breaking as a weak perturbation grows to nonlinear amplitudes, we
perform direct numerical simulation using Dedalus\cite{dedalus} a spectral
numerical method. Our region of interest is strictly interior to the WD and so
our simulation must strive to eliminate boundary effects resulting from
nonphysical boundaries. To do so, we adopt periodic boundary conditions in the
$x$ direction and damping layers (described in \autoref{ss:damping}) in the $z$
direction to damp perturbations exiting the domain. Correspondingly, we use a
Fourier basis in the $x$ direction and a Chebyshev in the $z$ direction.

@@

% Grid Size
% Viscosity + diffusivity, damping layers (to simulate radiative BC)
% timestepping? RK443, (CFL limiter??)

Our
simulation starts with all perturbation quantities at zero. We choose as
follows:
\begin{itemize}
    \item $k_{0x}$: Astrophysical IGWs have $k_{0x} \ll k_{0z}$, so we choose $k_{0x}
        = \frac{2\pi}{L_x}$ the smallest permitted wavenumber permitted by
        periodic boundary conditions.

    \item We choose $\omega_0$ by choosing it to produce a desired $k_{0z}$.
        Astrophysical systems generally also exhibit $k_{0z}H \gg 1$, where the
        stratification height is significantly larger than the vertical
        wavelength. However, in order for the simulation to both be well-damped
        at grid resolutions and be neglibily damped at leading order nonlinear
        wavelengths, we require $k_{0z} \ll \frac{1}{L_{NL}} \ll \frac{2\pi
        N_z}{L_z}$, where $L_{NL}$ denotes some characteristic length scale of
        nonlinear features. Finally, we need $L_z$ to be many $H$ to capture
        significant $e^{z/2H}$ growth and separate the linear and nonlinear
        amplitudes of $u_z$.

        Fixing $L_z = 10H$ to give $\sim e^4$ amplitude growth between the
        damping zones, this implies we require separation of scales $1 \ll
        k_{0z}H \ll \frac{H}{L_{NL}} \ll \frac{2\pi N_z}{10}$.

        Since $N_z$ is fixed by computational cost (we tried both $N_z = 512,
        N_z = 1024$), we choose $k_{0z} \sim \frac{2\pi}{H}$. This is less
        physically representative but gives more wavenumber space for the
        nonlinear cascade. We then invert $\omega_0 = \omega(k_{0x}, k_{0z})$.

    \item We choose $F$ forcing strength $1/20$ that of the nonlinear
        simulation, which keeps the flow amplitude sufficiently small to be
        treated in the linear approximation at all points in the simulation
        domain.
\end{itemize}

@@

To imitate an infinite fluid using a finite simulation domain, we use periodic
boundary conditions in the $x$ direction and damping layers at the top and
bottom of the $z$ direction. This damps waves that reach the edge of the
simulation domain without inducing nonphysical reflection. Specifically, these
are implemented by introducing linear damping terms to \autoref{se:nl_upsilon}:
\begin{align}
    \md{\Upsilon}{t} + u_z \pd{\ln \rho_0}{z} &=
        -\Gamma(z)\Upsilon,\nonumber\\
    \md{\vec{u}}{t} + \vec{\nabla}P + P\vec{\nabla}\Upsilon + g\hat{z} &=
        - \Gamma(z)\vec{u},\nonumber\\
    \frac{1}{2\tau}\s*{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}} &= \Gamma(z),\label{eq:Gamma}
\end{align}
where $z_B = 0.05L_z, z_T = 0.95L_z$ are the boundaries of the damping zone.
This strongly damps perturbations below $z_B$ and above $z_T$ with damping time
$\tau$, negligibly affects dynamics between $z_B, z_T$ and has transition width
governed by $\Delta z$; we use $\Delta z = 0.025L_z$. This prescription, used in
similar studies\cite{lecoanet_damp}, has the advantage of being smooth,
important for spectral methods. As these damping terms are linear, they are
retained in the linearized equations of motion as well.

The full implementation of the fluid equations in Dedalus is described in
\autoref{ss:strat_impl}.

\subsection{Linear Regime}\label{ss:lin}

We first test our numerical simulation with a ``linear'' simulation, where
all flow quantities are small: we assume  $\rho_1 \equiv \rho - \rho_0 \ll
\rho_0$ and $\vec{u} \cdot \vec{\nabla} \ll \partial_t$. We emphasize that our
``linear'' simulations solve the full nonlinear equations, just in the weak
perturbation limit.

$\omega$ is chosen by inverting $\omega(k_{0x}, k_{0z})$ dispersion relation for fixed
$k_{0x} = 2\pi / L_x$ and some desired $k_{0z} = -2\pi/H$, and $\sigma \lesssim
\frac{1}{k_{0z}}$ is used to excite a broad band of $k_z$ modes including the
desired $k_{0z}$ mode.

\subsection{Numerical Simulation}\label{ss:lin_ns}

A representative snapshot of this simulation
is provided in \autoref{fig:linear_nonu}. Key features to note:
\begin{itemize}
    \item $u_z$ grows with \autoref{eq:uz_lin} to reasonable accuracy.

    \item $S_{px}$ is seen to be constant, as would be expected by
        \autoref{eq:fpx_lin}.
\end{itemize}
We also include a qualitatively similar snapshot of the same simulation run
using the $\nu$ from the later nonlinear simulations in \autoref{fig:linear_nu}.
\begin{figure}[t]
    \centering
    \includegraphics[width=0.5\textwidth]{plots/linear_nonu.png}
    \caption{Snapshot of linear simulation after reaching steady state.
        Analytical predictions of $\abs*{u_z}, \bar{U}_0, S_{px}$ are shown in
        orange. Red dotted lines indicate the onset of the damping region and
        blue dotted lines denote the driving region.}\label{fig:linear_nonu}
\end{figure}
\begin{figure}[t]
    \centering
    \includegraphics[width=0.5\textwidth]{plots/linear_nu.png}
    \caption{Same as \autoref{fig:linear_nonu} but with $0.3\times $viscosity
    used for nonlinear simulations (YUBONOTE should be the same but that
    simulation mysteriously stopped without me noticing; restarting worked, so
    currently rerunning). Note the slight resulting dissipation in $S_{px}$.
    }\label{fig:linear_nu}
\end{figure}

@@

A
Navier-Stokes numerical viscosity $\nu$ is used to damp high wavenumbers and
regularize the nonlinear cascade at near grid resolution: $\nu \sim 0.2
\frac{\omega}{\abs*{k_{0z}}}\frac{L_z}{2\pi N_z}$ was found to be suitable for
$N_z = 1024$, where $k_{0z}$ is the wavenumber of the excited linear mode.
%%%

@@

\subsection{Numerical Simulation}

@@

Two representative snapshots from our simulation are provided in
\autoref{fig:nl} after the critical layer has had time to form. We may note that
the critical layer, where $S_{px}$ is absorbed and $\bar{U}_0 = \bar{U}_c$,
travels downwards as predicted.
\begin{figure}[t]
    \centering
    \begin{subfigure}{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/nl35_1.png}
        \caption{Earlier time snapshot. Legend is the same as
        \autoref{fig:linear_nonu} except $\bar{U}_c$ is marked in red on the
        third panel.}
    \end{subfigure}

    \begin{subfigure}{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/nl35_2.png}
        \caption{Later snapshot illustrating propagation of $z_c$.}
    \end{subfigure}
    \caption{Nonlinear numerical simulation.}\label{fig:nl}
\end{figure}

@@

For the simulation in \autoref{fig:nl}, we may define $z_c = \argmax_z
\pd{\bar{U}_0}{z}$. Computing $\pd{z_c}{t}$ using the analytical flux
\autoref{eq:fpx_lin} allows comparison to \autoref{eq:zc_anal}, which we exhibit
in \autoref{fig:zc_anal}. We see overall good agreement, though some small
deviation is expected since $S_{px}$ is not perfectly conserved owing to
numerical viscosity (and YUBONOTE $S_{px}$ is misestimated?).
\begin{figure}[t]
    \centering
    \includegraphics[width=0.5\textwidth]{plots/nl35_front_v.png}
    \caption{Comparison of simulated $\pd{z_c}{t}$ to analytical
    \autoref{eq:zc_anal}. Plot begins after formation of critical layer in
    simulation.}\label{fig:zc_anal}
\end{figure}

However, recalling \autoref{eq:crit_coeffs}, we only expect complete critical
layer absorption when $\mathrm{Ri} \gg \frac{1}{4}$. We may plot $\mathrm{Ri}$
over the same time interval in \autoref{fig:nl35_f_ri}. We may observe that the
Richardson number is initially decreasing but eventually levels out and
begins to increase again. This corresponds to an increasingly sharp critical
layer transition then subsequently a decreasingly sharp critical layer
transition.
\begin{figure}[t]
    \centering
    \includegraphics[width=0.5\textwidth]{plots/nl35_f_ri.png}
    \caption{Plot of $\mathrm{Ri}_{\max}(t) = \max_z
    \frac{N^2}{\bar{U}_0'}(t)$ over the simulation time.}\label{fig:nl35_f_ri}
\end{figure}

We argue that the Richardson number is bounded from below by viscosity. A
repeated simulation with a larger viscosity is shown in \autoref{fig:nl1_f_ri},
where the Richardson number does not go nearly as low.
\begin{figure}[t]
    \centering
    \includegraphics[width=0.5\textwidth]{plots/nl1_f_ri.png}
    \caption{Same as \autoref{fig:nl35_f_ri} but with $\sim 1.3\times$
    viscosity.}\label{fig:nl1_f_ri}
\end{figure}
